### 2022-02-08 chapters/onset/article.tex
updated ref for onset

%  PARAGRAPH DISEASES IN GENERAL
Coral diseases are a major threat to coral reef ecosystems and have led to significant declines in coral cover especially within the Caribbean region \citep{richardson1998coral, sutherland2004disease, aronson2001white, harvell2007coral, brandt2009dynamics}. One of the latest and the most damaging outbreak to date in Florida's Coral Reef (FCR) is the stony coral tissue loss disease (SCTLD) \citep{noaa2018}. First observed off the coast of Miami in 2014 by \cite{precht2016unprecedented}, the disease has since spread through the entire FCR \citep{muller2020spatial,dobbelaere2022} and has been observed in several territories of the Caribbean \citep{kramer2019map, meiling2021variable, estrada2021effects,heres2021ecological}. Although the causative agent of the disease remains unknown, hydrodynamics are likely to play an important role in its propagation as both modeling studies and ex situ experiments show evidence of waterborne disease transmission \citep{aeby2019pathogenesis,dobbelaere2020coupled,eaton2021measuring, meiling2021variable}. Furthermore, recent studies showed evidence that sediments can act as  a vector for the SCTLD \citep{rosales2020rhodobacterales, studivan2022reef}.

@@

While operating in a conventional way, dredged materials were pumped from the dredge to a spider barge and then transported to the US Environmental Protection Agency designated Ocean Dredge Material Disposal Site (ODMDS) located 4.7 nautical miles offshore. However, the suction mechanism was turned off during non-conventional rock-chopping activities in order to pre-treat very hard rock contained in the Anastasia and Fort Thompson formations between December 2013 and May 2014 \citep{miller2016detecting}. The Army Corps commissioned a report that provides a back-of-the-envelope estimating this practice could have resulted in up to 33 cm deposition over 874,121 m$^2$ of reef surrounding the outer entrance channel \citep{dobbelaere2020report}. Additionally, several studies reported that the impact of the dredging was widespread \citep{miller2016detecting}, causing the death of  $> 560,000$ corals within 0.5 km of the channel \citep{cunning2019extensive} and producing sediment plumes covering up to 11 km$^2$ of coral area within 5-10 km of the dredging \citep{barnes2015sediment}.

@@

The hydrodynamics of the entire FCR was modeled using the multi-scale ocean model SLIM\footnote{\url{ https://www.slim-ocean.be}}, which has already been extensively validated in the area \citep{frys20,dobbelaere2020coupled,dobbelaere2022}. SLIM uses an unstructured mesh whose resolution can be locally increased in order to accurately represent fine-scale flow features. The mesh used in this study was built following the same methodology as \cite{dobbelaere2022}, with a local refinement near PoM and in the Bay of Biscayne to achieve a resolution of 100 m in the vicinity of the dredged channel (Fig. \ref{fig:onset_mesh}A). It was made up of approximately $3.5\times 10^5$ triangles and was generated with the seamsh\footnote{\url{https://pypi.org/project/seamsh/}} Python library, which is based on the the open-source mesh generator GMSH \citep{geuzaine2009gmsh}. The model was run between October 15, 2013 and September 26, 2014 to cover the whole dredging period prior to the first observation of SCTLD by \cite{precht2016unprecedented}. Figure \ref{fig:onset_mesh}C depicts how a 100-m spatial resolution mesh simulated fine-scale details of the ocean currents, such as the flow acceleration between reefs and islands.

>>>>

%  PARAGRAPH DISEASES IN GENERAL
Coral diseases are a major threat to coral reef ecosystems and have led to significant declines in coral cover especially within the Caribbean region \citep{richardson1998coral, sutherland2004disease, aronson2001white, harvell2007coral, brandt2009dynamics}. One of the latest and the most damaging outbreak to date in Florida's Coral Reef (FCR) is the stony coral tissue loss disease (SCTLD) \citep{noaa2018}. First observed off the coast of Miami in 2014 by \cite{precht2016unprecedented}, the disease has since spread through the entire FCR \citep{muller2020spatial,dobbelaere2022connecting} and has been observed in several territories of the Caribbean \citep{kramer2019map, meiling2021variable, estrada2021effects,heres2021ecological}. Although the causative agent of the disease remains unknown, hydrodynamics are likely to play an important role in its propagation as both modeling studies and ex situ experiments show evidence of waterborne disease transmission \citep{aeby2019pathogenesis,dobbelaere2020coupled,eaton2021measuring, meiling2021variable}. Furthermore, recent studies showed evidence that sediments can act as  a vector for the SCTLD \citep{rosales2020rhodobacterales, studivan2022reef}.

@@

While operating in a conventional way, dredged materials were pumped from the dredge to a spider barge and then transported to the US Environmental Protection Agency designated Ocean Dredge Material Disposal Site (ODMDS) located 4.7 nautical miles offshore. However, the suction mechanism was turned off during non-conventional rock-chopping activities in order to pre-treat very hard rock contained in the Anastasia and Fort Thompson formations between December 2013 and May 2014 \citep{miller2016detecting}. The Army Corps commissioned a report that provides a back-of-the-envelope estimating this practice could have resulted in up to 33 cm deposition over 874,121 m$^2$ of reef surrounding the outer entrance channel \citep{usace2017}. Additionally, several studies reported that the impact of the dredging was widespread \citep{miller2016detecting}, causing the death of  $> 560,000$ corals within 0.5 km of the channel \citep{cunning2019extensive} and producing sediment plumes covering up to 11 km$^2$ of coral area within 5-10 km of the dredging \citep{barnes2015sediment}.

@@

The hydrodynamics of the entire FCR was modeled using the multi-scale ocean model SLIM\footnote{\url{ https://www.slim-ocean.be}}, which has already been extensively validated in the area \citep{frys20,dobbelaere2020coupled,dobbelaere2022connecting}. SLIM uses an unstructured mesh whose resolution can be locally increased in order to accurately represent fine-scale flow features. The mesh used in this study was built following the same methodology as \cite{dobbelaere2020coupled}, with a local refinement near PoM and in the Bay of Biscayne to achieve a resolution of 100 m in the vicinity of the dredged channel (Fig. \ref{fig:onset_mesh}A). It was made up of approximately $3.5\times 10^5$ triangles and was generated with the seamsh\footnote{\url{https://pypi.org/project/seamsh/}} Python library, which is based on the the open-source mesh generator GMSH \citep{geuzaine2009gmsh}. The model was run between October 15, 2013 and September 26, 2014 to cover the whole dredging period prior to the first observation of SCTLD by \cite{precht2016unprecedented}. Figure \ref{fig:onset_mesh}C depicts how a 100-m spatial resolution mesh simulated fine-scale details of the ocean currents, such as the flow acceleration between reefs and islands.
